Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-17 paper/paper.Rmd

  # institution:
  #   - name: Georgia Institute of Technology
  #     department: School of Electrical and Computer Engineering
  #     location: Atlanta, Georgia 30332--0250
  #     email: ece@datech.edu
  #     mark: 1
  #     author:
  #       - name: Michael Shell
  #   - name: Twentieth Century Fox
  #     location: Springfield, USA
  #     mark: 2
  #     author:
  #       - name: Homer Simpson
  #         email: homer@thesimpsons.com
  #   - name: Starfleet Academy
  #     location: San Francisco, California 96678-2391
  #     other: "Telephone: (800) 555--1212, Fax: (888) 555--1212"
  #     mark: 3
  #     author:
  #       - name: Montgomery Scott
  #   - name: Tyrell Inc.
  #     location: 123 Replicant Street, Los Angeles, USA 90210--4321
  #     mark: 4
  #     author:
  #       - name: Eldon Tyrell
  #         email: eldon@starfleet-academy.star
  #       - name: Roy Batty
  #         email: roy@replicant.offworld
keywords: ["keyword 1", "keyword 2"]
abstract: |
  Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata. The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point. There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals. Through this work we aim to close that gap by systematizing and extending existing knowledge. We first show that many of the existing methodologies can be collectively described by a generalized framework. We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level. Through simulation experiments involving various popular counterfactual generators and several benchmark datasets, we generate a total XX million Counterfactual Explanations and study the resulting domain and model shifts. We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in practice. Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches. Our simulation framework for studying recourse dynamics is fast and open-sourced.

# use some specific Tex packages if needed. 
# with_ifpdf: true
# with_cite: true
# amsmath need to be true to use with bookdown for referencing equations.
with_amsmath: true
# with_algorithmic: true
# with_array: true
# with_dblfloatfix: true

bibliography: https://raw.githubusercontent.com/pat-alt/bib/main/bib.bib
output: 
  bookdown::pdf_book:
    base_format: rticles::ieee_article
    includes:
      in_header: "tex/preamble.tex"
      before_body: "tex/before_body.tex"
#citation_sorting: none   ## used as sorting option of the biblatex package (if selected)
---

@@

{{< include sections/appendix.qmd >}}

>>>>

  # institution:
  #   - name: Georgia Institute of Technology
  #     department: School of Electrical and Computer Engineering
  #     location: Atlanta, Georgia 30332--0250
  #     email: ece@datech.edu
  #     mark: 1
  #     author:
  #       - name: Michael Shell
  #   - name: Twentieth Century Fox
  #     location: Springfield, USA
  #     mark: 2
  #     author:
  #       - name: Homer Simpson
  #         email: homer@thesimpsons.com
  #   - name: Starfleet Academy
  #     location: San Francisco, California 96678-2391
  #     other: "Telephone: (800) 555--1212, Fax: (888) 555--1212"
  #     mark: 3
  #     author:
  #       - name: Montgomery Scott
  #   - name: Tyrell Inc.
  #     location: 123 Replicant Street, Los Angeles, USA 90210--4321
  #     mark: 4
  #     author:
  #       - name: Eldon Tyrell
  #         email: eldon@starfleet-academy.star
  #       - name: Roy Batty
  #         email: roy@replicant.offworld
keywords: ["keyword 1", "keyword 2"]
abstract: |
  Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata. The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point. There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals. Through this work we aim to close that gap by systematizing and extending existing knowledge. We first show that many of the existing methodologies can be collectively described by a generalized framework. We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level. Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate a total of XX million counterfactuals and study the resulting domain and model shifts. We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in situations that involve competition for scarce resources. Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches. Our simulation framework for studying recourse dynamics is fast and open-sourced.

# use some specific Tex packages if needed. 
# with_ifpdf: true
with_cite: true
# amsmath need to be true to use with bookdown for referencing equations.
with_amsmath: true
# with_algorithmic: true
# with_array: true
# with_dblfloatfix: true

bibliography: https://raw.githubusercontent.com/pat-alt/bib/main/bib.bib
csl: ieee.csl
output: 
  bookdown::pdf_book:
    base_format: rticles::ieee_article
    includes:
      in_header: "tex/preamble.tex"
      before_body: "tex/before_body.tex"
link-citations: true
#citation_sorting: none   ## used as sorting option of the biblatex package (if selected)
---

@@

### 2022-08-17 paper/sections/conclusion.rmd

This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse. We demonstrate that long-held beliefs as to what defines optimality in AR, may not be applicable in situations that involve decisions over scarce resources. Specifically, we run experiments that simulate the application of recourse in practice using various state-of-the-art counterfactual generators and find that all of them induce substantial domain and model shifts. We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse. By proposing an adapted counterfactual search objective that incorporates this cost, we make that paradigm shift explicit. We show that this modified objective lends itself to mitigation strategies that can be used to effectively decrease the magnitude of induced domain and model shifts. Through our work we hope to inspire future research on this important topic. To this end we have open-sourced all of our code along with a Julia package - [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md). The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022CounterfactualExplanations]. That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.

>>>>

This work has revisited and extended some of the most general and defining concepts underlying the literature on Counterfactual Explanations and, in particular, Algorithmic Recourse. We demonstrate that long-held beliefs as to what defines optimality in AR, may not be suitable in contexts that involves large groups of individuals facing adverse outcomes. Specifically, we run experiments that simulate the application of recourse in practice using various state-of-the-art counterfactual generators and find that all of them induce substantial domain and model shifts. We argue that these shifts should be considered as an expected external cost of individual recourse and call for a paradigm shift from individual to collective recourse in these types of situations. By proposing an adapted counterfactual search objective that incorporates this cost, we make that paradigm shift explicit. We show that this modified objective lends itself to mitigation strategies that can be used to effectively decrease the magnitude of induced domain and model shifts. Through our work we hope to inspire future research on this important topic. To this end we have open-sourced all of our code along with a Julia package - [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md). The package is built on top of `CounterfactualExplanations.jl` and inherits its extensibility [@altmeyer2022CounterfactualExplanations]. That is to say that future researchers should find it relatively easy to replicate, modify and extend the simulation experiments presented here and apply to their own custom counterfactual generators.

### 2022-08-17 paper/sections/discussion.rmd

Our results in Section \@ref(empirical-2) indicate that state-of-the-art approaches to Algorithmic Recourse induce substantial domain and model shift, if implemented at scale in practice. These induced shifts can and should be considered as an (expected) external cost of individual recourse. While they do not affect the individual directly as long as we look at the individual in isolation, they can been seen to affect the broader group of stakeholders in automated data-driven decision-making. We have seen, for example, that out-of-sample model performance generally deteriorates in our simulation experiments. In practice, this can be seen as a cost to model owners, that is the group of stakeholders using the model as decision-making tool. As we have set out in the introduction, these model owners will generally be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether. This in turn is costly to those individuals that would otherwise derive utility from being offered recourse.

>>>>

Our results in Section \@ref(empirical-2) indicate that state-of-the-art approaches to Algorithmic Recourse induce substantial domain and model shift, if implemented at scale in practice. These induced shifts can and should be considered as an (expected) external cost of individual recourse. While they do not affect the individual directly as long as we look at the individual in isolation, they can been seen to affect the broader group of stakeholders in automated data-driven decision-making. We have seen, for example, that out-of-sample model performance generally deteriorates in our simulation experiments. In practice, this can be seen as a cost to model owners, that is the group of stakeholders using the model as decision-making tool. As we have set out in Example \@ref(exm:student) of our introduction, these model owners may be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether. This in turn is costly to those individuals that would otherwise derive utility from being offered recourse.

### 2022-08-17 paper/sections/introduction.rmd

```{r poc, fig.cap="Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d)."}
knitr::include_graphics("www/poc.png")
```

>>>>

```{r poc, fig.cap="Dynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class ($y=0$) are marked in blue and samples of the positive class ($y=1$) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class."}
knitr::include_graphics("www/poc.png")
```

### 2022-08-17 paper/sections/mitigation.rmd

\begin{equation}
\begin{aligned}
\text{extcost}(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime) (\#eq:clap)
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
\text{extcost}(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})  (\#eq:grav)
\end{aligned}
\end{equation}

@@

```{r mitigation, fig.cap="Illustrative example demonstrating the properties of the various mitigation strategies."}
knitr::include_graphics("www/mitigation.png")
```

>>>>

\begin{equation}
\begin{aligned}
\text{extcost}(f(\mathbf{s}^\prime)) = l(M(f(\mathbf{s}^\prime)),y^\prime) (\#eq:clap)
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
\text{extcost}(f(\mathbf{s}^\prime)) = \text{dist}(f(\mathbf{s}^\prime),\bar{x})  (\#eq:grav)
\end{aligned}
\end{equation}

@@

```{r mitigation, fig.cap="Illustrative example demonstrating the properties of the various mitigation strategies. Samples from the negative class ($y=0$) are marked in blue while samples of the positive class ($y=1$) are marked in orange."}
knitr::include_graphics("www/mitigation.png")
```

### 2022-08-17 paper/sections/related.rmd

A framework for Counterfactual Explanations was first proposed in 2017 by @wachter2017counterfactual and has served as the baseline for most methodologies that have been proposed since then. Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$. Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $H=h(x\prime)$ incurred by individual $x$ when moving to a counterfactual state $x\prime$ such that the predicted outcome $M(x\prime)$ corresponds to some target outcome $y^*$:

\begin{equation}
\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* (\#eq:obj)
\end{equation}

@@

\begin{equation}
x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime) (\#eq:solution)
\end{equation}

>>>>

A framework for Counterfactual Explanations was first proposed in 2017 by @wachter2017counterfactual and has served as the baseline for most methodologies that have been proposed since then. Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$. Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $C=\text{cost}(x\prime)$ incurred by individual $x$ when moving to a counterfactual state $x\prime$ such that the predicted outcome $M(x\prime)$ corresponds to some target outcome $y^*$:

\begin{equation}
\min_{x\prime \in \mathcal{X}} \text{cost}(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* (\#eq:obj)
\end{equation}

@@

\begin{equation}
x\prime = \arg \min_{x\prime}  \text{yloss}(M(x\prime),y^*) + \lambda \text{cost}(x\prime) (\#eq:solution)
\end{equation}
